---
title: "Figure 4"
author: "Oliver Polzer"
date: "2025-05-27"
output: html_document
---

```{r setup, include=FALSE}
library(rmarkdown)
library(tinytex)
library(knitr)
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, cache = TRUE)
```

### 1. Background

The following figure was carried out based on the following publications  
1.[Ibrayeva et al., 2021](https://doi.org/10.1016/j.stem.2021.03.018)  
2.[Harris et al., 2021](https://doi.org/10.1016/j.stem.2021.01.003)    
[Generation of the Seurat object from Harris et al.](https://github.com/harrislachlan/lifelong_stemcells/tree/main/markdown)    
2.[Wu et al., 2025](https://doi.org/10.1038/s41593-024-01848-4)  
[Seurat object from Wu et al.](https://github.com/JessbergerLab/AgingNeurogenesis_Transcriptomics)  

In order to generate these figures the seurat object from Harris et al. needs to be generated (according to original markdown document, see link above) and the seurat object from Wu et al. can be directly downloaded from the link above.  

This reproducible and dynamic report was created using Rmarkdown and the [Knitr package](https://yihui.name/knitr/), and summarizes the basic code and outputs (plots, tables, etc).  



***
### Figure 4 A

GO terms from the original studies were extracted. The extracted GO terms from yound and old NSCs were then summarized using [REVIGO](http://revigo.irb.hr/)

***
### Figure 4 B

```{r SenMayo, eval = FALSE}
# Load whole Harris dataset----
#Please follow the steps from the original paper, once object is generated subsequent steps can be performed
load("nsc.RData")

Idents(nsc.integrated) <- nsc.integrated$quiescence
Idents(nsc.integrated) <- nsc.integrated$Age
DimPlot(nsc.integrated, group.by = "Age", pt.size = 1.5, order = c("6mo", "2mo", "1mo"), cols = c("yellow2", "orange2", "purple4")) %>%
  + theme_minimal() %>% 
  + NoLegend() + labs(title = "")

#Assign ModuleScore
# Load the senescence gene lists
senMayo <- read.csv("SenMayo.csv", header = FALSE)

# Convert data frames to vectors
senMayo <- as.vector(senMayo$V1)

# Define the signatures list
signatures <- list(
  SenMayoList = senMayo
)

# Calculate module scores using AddModuleScore_UCell
nsc.integrated <- AddModuleScore_UCell(nsc.integrated, features = signatures, name = "Senescence_Score")

seurat.object <- AddModuleScore_UCell(nsc.integrated, features = signatures, name = NULL)

#Visualize module scores using FeaturePlot
FeaturePlot(seurat.object, pt.size = 1, min.cutoff = 0.04 , reduction = "umap", features = names(signatures))

```



***
### Figure 4 C and Supplementary figure 2

```{r Cellchat, eval = FALSE}
library(CellChat)
library(Seurat)
library(dplyr)
library(gridExtra)
library(ComplexHeatmap)

#Load Wu seurat object
seurat_object <- readRDS("C:/Users/LucassenEDU/Desktop/Wu et al/Wu_analysis/seurat_Chromium_all.rds")
Idents(seurat_object) <- "Celltype"


#Split by age and create CellChat object (Skip this if you already have the objects)----

# Step 1: Split Seurat object by Age
seurat_young <- subset(seurat_object, subset = Age == "Young")
seurat_middle <- subset(seurat_object, subset = Age == "Middle")
seurat_old <- subset(seurat_object, subset = Age == "Old")

# Step 2: Create CellChat objects for each age group
cellchat_young <- createCellChat(object = seurat_young, group.by = "ident", assay = "RNA")
cellchat_middle <- createCellChat(object = seurat_middle, group.by = "ident", assay = "RNA")
cellchat_old <- createCellChat(object = seurat_old, group.by = "ident", assay = "RNA")

# Step 3: Load and assign the CellChat database
CellChatDB <- CellChatDB.mouse  
CellChatDB.use <- subsetDB(CellChatDB)  

cellchat_young@DB <- CellChatDB.use
cellchat_middle@DB <- CellChatDB.use
cellchat_old@DB <- CellChatDB.use

# Step 4: Subset expression data to reduce computation
cellchat_young <- subsetData(cellchat_young)
cellchat_middle <- subsetData(cellchat_middle)
cellchat_old <- subsetData(cellchat_old)

# Step 5: Identify overexpressed genes and interactions
cellchat_young <- identifyOverExpressedGenes(cellchat_young)
cellchat_young <- identifyOverExpressedInteractions(cellchat_young)

cellchat_middle <- identifyOverExpressedGenes(cellchat_middle)
cellchat_middle <- identifyOverExpressedInteractions(cellchat_middle)

cellchat_old <- identifyOverExpressedGenes(cellchat_old)
cellchat_old <- identifyOverExpressedInteractions(cellchat_old)

# Step 6: Compute communication probability
cellchat_young <- computeCommunProb(cellchat_young, type = "triMean")
cellchat_middle <- computeCommunProb(cellchat_middle, type = "triMean")
cellchat_old <- computeCommunProb(cellchat_old, type = "triMean")

# Step 7: Compute the cell-cell communication at a signaling pathway level
cellchat_young <- computeCommunProbPathway(cellchat_young)
cellchat_middle <- computeCommunProbPathway(cellchat_middle)
cellchat_old <- computeCommunProbPathway(cellchat_old)

cellchat_young <- aggregateNet(cellchat_young)
cellchat_middle <- aggregateNet(cellchat_middle)
cellchat_old <- aggregateNet(cellchat_old)

saveRDS(cellchat_young, file = "CellChat_Wu_Young.rds")
saveRDS(cellchat_middle, file = "CellChat_Wu_Middle.rds")
saveRDS(cellchat_old, file = "CellChat_Wu_Old.rds")


#Only look at pairwise comparison between Young and Old----

# Define the list of CellChat objects
cellchat_list_pairwise <- list(Young = cellchat_young, Old = cellchat_old)
cellchat_merged_pairwise <- mergeCellChat(cellchat_list_pairwise, add.names = names(cellchat_list_pairwise))

# Compare total number of interactions and interaction strength
gg1 <- compareInteractions(cellchat_merged_pairwise, show.legend = FALSE, group = c(1,2))
gg2 <- compareInteractions(cellchat_merged_pairwise, show.legend = FALSE, group = c(1,2), measure = "weight")
plot1 <- gg1 + gg2
ggsave("2025-03-13_total_interactions_comparison.svg", plot1)

# Compare number of interactions and interaction strength among different cell populations
svg(filename = "2025-03-13_Interaction_network.svg", width = 12, height = 6)
par(mfrow = c(1,2), xpd=TRUE)
netVisual_diffInteraction(cellchat_merged_pairwise, weight.scale = TRUE)
dev.off()


# Heatmap showing differential interactions among cell populations
gg3 <- netVisual_heatmap(cellchat_merged_pairwise)
gg4 <- netVisual_heatmap(cellchat_merged_pairwise, measure = "weight")

# Save heatmaps as SVG
svg("2025-03-13_heatmap_interactions.svg", width = 6, height = 4)
draw(gg3)
draw(gg4)
dev.off()

# Identify cell populations with significant changes in signaling
num.link <- sapply(cellchat_list_pairwise, function(x) {rowSums(x@net$count) + colSums(x@net$count) - diag(x@net$count)})
weight.MinMax <- c(min(num.link), max(num.link))

gg <- list()
for (i in 1:length(cellchat_list_pairwise)) {
  gg[[i]] <- netAnalysis_signalingRole_scatter(cellchat_list_pairwise[[i]], title = names(cellchat_list_pairwise)[i], weight.MinMax = weight.MinMax)
}
plot3 <- patchwork::wrap_plots(plots = gg)
ggsave("2025-03-13_signaling_role_analysis.svg", plot3)

# Identify signaling changes of specific cell populations
gg6 <- netAnalysis_signalingChanges_scatter(cellchat_merged_pairwise, idents.use = "qNSC")
gg7 <- netAnalysis_signalingChanges_scatter(cellchat_merged_pairwise, idents.use = "nIPC")
gg8 <- netAnalysis_signalingChanges_scatter(cellchat_merged_pairwise, idents.use = "Neuroblast")
plot4 <- patchwork::wrap_plots(plots = list(gg6, gg7, gg8))
ggsave("2025-03-13_signaling_changes_qNSC.svg", gg6)
ggsave("2025-03-13_signaling_changes_nIPC.svg", gg7)
ggsave("2025-03-13_signaling_changes_neuroblast.svg", gg8)


```
