---
title: "Figure 3"
author: "Oliver Polzer"
date: "2025-05-27"
output: html_document
---
```{r setup, include=FALSE}
library(rmarkdown)
library(tinytex)
library(knitr)
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, cache = TRUE)
```

### 1. Background

The following analysis was carried out based on two publicly available datasets  
1.[Harris et al., 2021](https://doi.org/10.1016/j.stem.2021.01.003)  
[Generation of the Seurat object + pseudotime values from Harris et al.](https://github.com/harrislachlan/lifelong_stemcells/tree/main/markdown)  
2.[Shin et al., 2015](https://doi.org/10.1016/j.stem.2015.07.013)   

In order to generate these figures the seurat object from Harris et al. needs to be generated (according to original markdown document, see link above).  

This reproducible and dynamic report was created using Rmarkdown and the [Knitr package](https://yihui.name/knitr/), and summarizes the basic code and outputs (plots, tables, etc).  



***
### Figure 3 A - D

```{r Pseudotime Shin Harris, eval = FALSE}
##Load packages----
library(readxl)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(ggpubr)
library(ggdist)
library(ggforce)
library(xtable)
library(ggthemes)

# Specify the path to your Excel file of both pseudotime trajectories----
# Pseudotime values from Shin et al. can be found in the original paper or in the GitHub repository
# Pseudotime values from Harris et al. need to be generated using the pipeline from the original paper
excel_file_shin <- "YourDirectory/Shin_Pseudotime_all_genes.xlsx"
excel_file_harris <- "YourDirectory/Harris_Pseudotime_all_genes.csv"

# Read the Excel file into a data frame----
data_shin <- read_excel(excel_file_shin)
colnames(data_shin)[1] <- "GeneName"
data_shin_df <- data.frame(data_shin) 

data_shin_df2 <- data_shin_df[,-1]
rownames(data_shin_df2) <- data_shin_df[,1]

data_harris <- read.csv(excel_file_harris)
colnames(data_harris)[1] <- "GeneName"
data_harris_df <- data.frame(data_harris)

data_harris_df2 <- data_harris_df[,-1]
rownames(data_harris_df2) <- data_harris_df[,1]
# Remove the row with the name "age"
data_harris_df2 <- data_harris_df2[rownames(data_harris_df2) != "quiescence", ]
data_harris_df2 <- data_harris_df2[rownames(data_harris_df2) != "...1", ]


# all gene lists of NSC populations from selected studies ----

# Define the studies you want to process and their corresponding gene list files----
# Example here are all population with NSC characteristics (Figure 3 A and C)
# For Figure 3 B and D please use the respective populations

studies <- c("Artegiani_NSC","Batiuk_AST4","Bielefeld_A1", "Harris_NSC","Hochgerner_RGL", "Wu_qNSC","Shin_NSC", "Overlap_NSC")
files <- paste0(studies, ".txt")  # assuming files are named like "Shin_NSC.txt", etc.

# Read gene lists into a named list
gene_lists <- setNames(lapply(files, readLines), studies)

# Helper function to filter data based on a gene list
filter_data <- function(data_df, gene_list) {
  df <- data_df[rownames(data_df) %in% gene_list, ]
  # Convert character columns to numeric if needed
  df <- dplyr::mutate_if(df, is.character, as.numeric)
  return(df)
}

# For each study, filter both the Shin and Harris datasets,
# compute column means, combine with x-values, and z-score normalize
process_study <- function(study, data_shin, data_harris, x_shin, x_harris) {
  # Filter data for the given gene list
  fs <- filter_data(data_shin, gene_lists[[study]])
  fh <- filter_data(data_harris, gene_lists[[study]])
  
  # Calculate column means
  mean_s <- colMeans(fs)
  mean_h <- colMeans(fh)
  
  # Combine with x-values (assuming Pseudotime is in these data frames)
  plot_s <- cbind(x_shin, mean = mean_s)
  plot_h <- cbind(x_harris, mean = mean_h)
  
  # Z-score normalization
  plot_s$zscore <- as.numeric(scale(plot_s$mean))
  plot_h$zscore <- as.numeric(scale(plot_h$mean))
  
  # Add a dataset column (prefixing with "s_" for Shin)
  plot_s$dataset <- paste0("s_", study)
  plot_h$dataset <- paste0("h_", study)
  
  # Return both datasets as a list 
  return(list(shin = plot_s, harris = plot_h))
}

# Create x-values from your data (modify row index as appropriate)
x_values_s <- as.data.frame(t(data_shin_df2[1, ]))
x_values_h <- as.data.frame(t(data_harris_df2[1, ])) %>% 
  dplyr::mutate_if(is.character, as.numeric)

# Process each study and combine Shin and Harris results
result_list <- lapply(studies, process_study, data_shin = data_shin_df2, 
                      data_harris = data_harris_df2, 
                      x_shin = x_values_s, x_harris = x_values_h)
names(result_list) <- studies


# Extract and combine the Shin datasets
plot_data_s_combined <- do.call(rbind, lapply(result_list, function(x) {
  # Keep only relevant columns and assume "Pseudotime" is a column in x_shin
  x$shin[, c("Pseudotime","mean", "zscore", "dataset")]
}))

# Extract and combine the Harris datasets
plot_data_h_combined <- do.call(rbind, lapply(result_list, function(x) {
  x$harris[, c("Lineage1","mean", "zscore", "dataset")]
}))

#Plot graphs----

# Generate the plot for Shin dataset
s_combined_plot <- ggplot(plot_data_s_combined, aes(x = Pseudotime, y = plot_data_s_combined$zscore, color = dataset)) +
  geom_point(size = 1, alpha = 0.2) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(
    x = "Pseudotime",
    y = "Mean gene expression levels",
    title = "NSC Gene Expression Across Pseudotime"
  ) +
  theme_tufte(base_family = "Arial") +
  theme(
    plot.margin = margin(1, 1, 1, 1, "cm"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line = element_line(colour = "black"),
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 14),
    legend.position = "bottom",
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14)
  ) +
  xlim(0, 1) +
  ylim(-2, 2)

print(s_combined_plot)


# Generate the plot for Harris dataset
h_combined_plot <- ggplot(plot_data_h_combined, aes(x = Lineage1, y = plot_data_h_combined$zscore, color = dataset)) +
  geom_point(size = 1, alpha = 0.2) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(
    x = "Pseudotime",
    y = "Mean gene expression levels",
    title = "NSC Gene Expression Across Pseudotime (Harris)"
  ) +
  theme_tufte(base_family = "Arial") +
  theme(
    plot.margin = margin(1, 1, 1, 1, "cm"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line = element_line(colour = "black"),
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 14),
    legend.position = "bottom",
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14)
  ) +
  xlim(0, 20) +
  ylim(-1, 5)

# Print the plot
print(h_combined_plot)

```

***
### Figure 3 E and F

```{r Pseudotime Harris, eval = FALSE}

# Load packages----

library(dplyr)
library(tidyr)
library(ggplot2)
library(readxl)


# Read the Shin Excel File and Prepare the Data
excel_file_shin <- "YourDirectory/Shin_Pseudotime_all_genes.xlsx"

# Read the Excel file; assumes first column contains gene names
data_shin <- read_excel(excel_file_shin)
data_shin_df <- data.frame(data_shin) %>%
  rename(GeneName = 1)

# Remove the first column (now stored in GeneName) and set row names to gene names
data_shin_df2 <- data_shin_df[,-1]
rownames(data_shin_df2) <- data_shin_df[,1]


# Extract Pseudotime and Gene Expression Data----

# Assume one row is named "Pseudotime" (with pseudotime values for each sample)
# Extract the pseudotime vector and remove that row from the data
pseudotime <- as.numeric(data_shin_df2["Pseudotime", ])
data_shin_genes <- data_shin_df2[rownames(data_shin_df2) != "Pseudotime", ]


# Convert Wide Data to Long Format

# Before pivoting, create a data frame mapping Sample names to Pseudotime.
# Here, column names (except "Pseudotime") are assumed to be the sample IDs.
pseudotime_df <- data.frame(
  Sample = colnames(data_shin_df2)[colnames(data_shin_df2) != "Pseudotime"],
  Pseudotime = pseudotime,
  stringsAsFactors = FALSE
)

# Add a column "Gene" to the gene expression data using row names
data_shin_genes$Gene <- rownames(data_shin_genes)

# Pivot from wide to long format: each row is a gene-sample combination
data_shin_long <- data_shin_genes %>%
  pivot_longer(
    cols = -Gene,            # all columns except "Gene"
    names_to = "Sample",     # new column for sample IDs
    values_to = "Expression" # new column for expression values
  )

# Join the pseudotime information by matching Sample names
data_shin_long <- left_join(data_shin_long, pseudotime_df, by = "Sample")


# Filter for Selected Genes
# add here the genes of interest
selected_genes <- c("Apoe", "Ccnd2", "Ascl1", "Cdk6", "Ezh2", "Hmgb2", "Insm1", "Lmnb1", "Neurog2")

data_shin_long_filtered <- data_shin_long %>%
  filter(Gene %in% selected_genes)

# Ensure the Expression column is numeric (in case it was read as character)
data_shin_long_filtered$Expression <- as.numeric(data_shin_long_filtered$Expression)

# Z-score Normalize the Expression Data

# Z-score normalization of gene expression
data_shin_long_filtered <- data_shin_long_filtered %>%
  group_by(Gene) %>%
  mutate(
    ZScore_Expression = (Expression - mean(Expression, na.rm = TRUE)) / sd(Expression, na.rm = TRUE)
  ) %>%
  ungroup()


# Plot Gene Expression vs. Pseudotime

# Option A: All genes in one plot, colored by gene
p_shin_all <- ggplot(data_shin_long_filtered, aes(x = Pseudotime, y = ZScore_Expression, color = Gene)) +
  geom_smooth(method = "loess", se = TRUE, aes(fill = Gene), alpha = 0.3) +  # Shading for 95% CI
  labs(
    title = "Expression of Selected Genes vs. Pseudotime (Shin)",
    x = "Pseudotime",
    y = "Expression"
  ) +
  theme_minimal() +
  theme(legend.position = "none") +  # Optionally remove legend if too many genes
  facet_wrap(~ Gene, scales = "free_y") +  # Create a separate panel for each gene
  coord_cartesian(ylim = c(-2, 2))  # Set y-axis limits from -2 to 2

print(p_shin_all)

# Option B: Faceted plot (one panel per gene)
p_shin_facet <- ggplot(data_shin_long_filtered, aes(x = Pseudotime, y = Expression)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_smooth(method = "loess", se = FALSE, color = "red") +
  facet_wrap(~ Gene, scales = "free_y") +
  labs(
    title = "Expression of Selected Genes vs. Pseudotime (Faceted)",
    x = "Pseudotime",
    y = "Expression"
  ) +
  theme_minimal()

print(p_shin_facet)


###For NSC signature##
# 1. Load the gene list from the file
selected_genes <- readLines("Overlap_NP.txt")

# 2. Filter the data for the selected genes
data_shin_long_filtered <- data_shin_long %>%
  filter(Gene %in% selected_genes)

# Ensure the Expression column is numeric (in case it was read as character)
data_shin_long_filtered$Expression <- as.numeric(data_shin_long_filtered$Expression)

# 3. Z-score normalize the Expression Data
data_shin_long_filtered <- data_shin_long_filtered %>%
  group_by(Gene) %>%
  mutate(
    ZScore_Expression = (Expression - mean(Expression, na.rm = TRUE)) / sd(Expression, na.rm = TRUE)
  ) %>%
  ungroup()

# 4. Calculate the average Z-score expression for the selected genes at each pseudotime point
data_avg_expression <- data_shin_long_filtered %>%
  group_by(Pseudotime, Sample) %>%
  summarise(
    Avg_ZScore_Expression = mean(ZScore_Expression, na.rm = TRUE)
  )

# 5. Plot the average expression (smoothed curve) for the selected genes
p_shin_all_avg <- ggplot(data_avg_expression, aes(x = Pseudotime, y = Avg_ZScore_Expression)) +
  geom_smooth(method = "loess", se = TRUE, aes(fill = "Average"), alpha = 0.3) +  # Shading for 95% CI
  labs(
    title = "Average Expression of Selected Genes vs. Pseudotime (Shin)",
    x = "Pseudotime",
    y = "Average Z-Score Expression"
  ) +
  theme_minimal() +
  coord_cartesian(ylim = c(-2, 2))  # Set y-axis limits from -2 to 2

# Print the plot
print(p_shin_all_avg)

ggsave("2025-03-09_Pseudotime_NPsignature_plot_shin_all.svg", plot = p_shin_all_avg, device = "svg", width = 10, height = 8)




```

***
### Figure 3 G

```{r Geneoverlap with Wu}

#Load necessary libraries----
library(ggplot2)
library(dplyr)
library(tidyr)

# Define the dataset
# Modules were extracted from Wu et al. and gene overlap was determined in Excel
data <- data.frame(
  Module = c("Module 1", "Module 2", "Module 3", "Module 4", "Not Included"),
  NSC = c(49, 9, 0, 0, 0),  # NSC genes in each module + not included
  NP = c(0, 2, 57, 0, 3),   # NP genes in each module + not included
  NB = c(0, 0, 17, 61, 0)   # NB genes in each module + not included
)

# Define total number of genes per signature
total_NSC <- sum(data$NSC)
total_NP <- sum(data$NP)
total_NB <- sum(data$NB)

# Convert data to long format
data_long <- data %>%
  pivot_longer(cols = NSC:NB, names_to = "Signature", values_to = "Count")

# Add total genes per signature and compute percentages
data_long <- data_long %>%
  mutate(Total_Signature = case_when(
    Signature == "NSC" ~ total_NSC,
    Signature == "NP" ~ total_NP,
    Signature == "NB" ~ total_NB
  )) %>%
  mutate(Percentage = (Count / Total_Signature) * 100)

# Separate "Not Included" category
not_included <- data_long %>% filter(Module == "Not Included")
data_long <- data_long %>% filter(Module != "Not Included")

# Merge with "Not Included" to ensure each bar sums to 100%
data_long <- data_long %>%
  left_join(not_included %>% select(Signature, NotIncluded_Percentage = Percentage), by = "Signature") %>%
  mutate(NotIncluded_Percentage = ifelse(is.na(NotIncluded_Percentage), 0, NotIncluded_Percentage)) %>%
  mutate(Present_Percentage = Percentage, # Renaming for clarity
         Remaining_Percentage = 100 - Present_Percentage) %>%
  select(Module, Signature, Present_Percentage, Remaining_Percentage)

# Convert back to long format for plotting
data_long <- data_long %>%
  pivot_longer(cols = c(Present_Percentage, Remaining_Percentage),
               names_to = "Category", values_to = "Percentage")

# Ensure Signature order (NSC → NP → NB)
data_long$Signature <- factor(data_long$Signature, levels = c("NSC", "NP", "NB"))

# Ensure "Present_Percentage" is plotted first (on the bottom)
data_long$Category <- factor(data_long$Category, levels = c("Present_Percentage", "Remaining_Percentage"))

# Create a stacked bar plot where each signature sums to 100%, and signatures are side by side
p <- ggplot(data_long, aes(x = Module, y = Percentage, fill = Category)) +
  geom_bar(stat = "identity", position = "stack") +  # Stacked bars
  facet_wrap(~ Signature, scales = "free_x") +  # Panel by Signature (NSC, NP, NB)
  labs(title = "Signature Distribution Within Each Module (100% per Signature)",
       x = "Module",
       y = "Percentage (%)") +
  theme_minimal() +
  scale_fill_manual(values = c("Remaining_Percentage" = "gray", "Present_Percentage" = "blue"),
                    labels = c("Not Included", "Present")) +  # Fix legend labels
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("2025-02-27_GeneOverlap_modules_wuplot.svg", plot = p, device = "svg", width = 6, height = 4)

```

