---
title: "Figure 1"
author: "Oliver Polzer"
date: "2025-05-27"
output: 
    html_document: 
    toc: true
    Toc_float: true
---
```{r setup, include=FALSE}
library(rmarkdown)
library(tinytex)
library(knitr)
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, cache = TRUE)
```

### 1. Background

The following analysis was carried out based on publicly available datasets  
1.[Shin et al., 2015](https://doi.org/10.1016/j.stem.2015.07.013)   
2.[Artegiani et al., 2017](https://doi.org/10.1016/j.celrep.2017.11.050)  
3.[Hochgerner er al., 2018](https://doi.org/10.1038/s41593-017-0056-2)  
4.[Batiuk et al., 2020](https://doi.org/10.1038/s41467-019-14198-8)  
5.[Harris et al., 2021](https://doi.org/10.1016/j.stem.2021.01.003)  
6.[Bielefeld et al., 2024 ](https://doi.org/10.1101/2023.03.04.531101)  
7.[Wu et al., 2025](https://doi.org/10.1038/s41593-024-01848-4)  
This reproducible and dynamic report was created using Rmarkdown and the [Knitr package](https://yihui.name/knitr/), and summarizes the basic code and outputs (plots, tables, etc).  

Information regarding the separated studies can be found in Table 2 of the paper.


***


### 2. Figure 1

#### Gene lists used to identify/characterize neural stem cells or neural progenitor cells were extracted from the studies above and used for hierarchical clustering using the hclust function based on the similiarity of genes across studies.

```{r clustering}

# Load necessary packages----
library(dplyr)
library(VennDiagram)
library(gplots)
library(dendextend)
library(UpSetR)
library(plotly)
library(gridExtra)
library("ggvenn")

# For Figure 1 A
# Read and store genes----
# Set the path to the directory where your gene list files of are located
# All gene lists can be found in the folder GeneLists in GitHub
gene_list_dir <- "C:/Users/LucassenEDU/Desktop/R_code_rebuttal/GeneLists_NSC_NPC"

# List all files in the directory
gene_list_files <- list.files(gene_list_dir, pattern = ".txt", full.names = TRUE)

# Read and store gene lists in a list of data frames
gene_lists <- lapply(gene_list_files, function(file) {
  # Extract the file name without extension
  new_name <- tools::file_path_sans_ext(basename(file))
  
  # Read the gene list from the file
  gene_list <- read.delim(file, header = FALSE, stringsAsFactors = FALSE)
  
  # Rename the gene list with the new name
  names(gene_list) <- new_name
  
  # Return the gene list
  gene_list
})

# Perform pairwise comparisons and store the results----
pairwise_results <- list()

for (i in 1:(length(gene_lists) - 1)) {
  for (j in (i+1):length(gene_lists)) {
    set1 <- gene_lists[[i]][[1]]
    set2 <- gene_lists[[j]][[1]]
    common_genes <- intersect(set1, set2)
    pairwise_results[[paste0("Gene List", i, "_Gene List", j)]] <- common_genes
  }
}

# Quantification of Similarity
similarity_matrix <- matrix(0, nrow = length(gene_lists), ncol = length(gene_lists))

for (i in 1:(length(gene_lists) - 1)) {
  for (j in (i+1):length(gene_lists)) {
    set1 <- gene_lists[[i]][[1]]
    set2 <- gene_lists[[j]][[1]]
    common_genes <- intersect(set1, set2)
    similarity_matrix[i, j] <- length(common_genes)
    similarity_matrix[j, i] <- length(common_genes)
  }
}

# Print similarity matrix
rownames(similarity_matrix) <- c("Artegiani_NP", "Artegiani_NSC", 
                                 "Batiuk_AST4",
                                  "Bielefeld_A1",
                                 "Bielefeld_NSC1","Bielefeld_NSC2",
                                "Bielefeld_RGL",
                                 "Harris_Nb",
                                 "Harris_NSC","Harris_NSC_IPCs",
                                 "Hochgerner_Nb",
                                 "Hochgerner_nIPC","Hochgerner_RGL",
                                 "Shin_NSC",
                                 "Wu_Nb",
                                 "Wu_nIPC", "Wu_qNSC")

colnames(similarity_matrix) <- c("Artegiani_NP", "Artegiani_NSC", 
                                 "Batiuk_AST4",
                                  "Bielefeld_A1",
                                 "Bielefeld_NSC1","Bielefeld_NSC2",
                                  "Bielefeld_RGL",
                                 "Harris_Nb",
                                 "Harris_NSC","Harris_NSC_IPCs",
                                 "Hochgerner_Nb",
                                 "Hochgerner_nIPC","Hochgerner_RGL",
                                 "Shin_NSC",
                                 "Wu_Nb",
                                 "Wu_nIPC", "Wu_qNSC")


# Create a dendrogram using hierarchical clustering
dendrogram <- hclust(as.dist(1 - (similarity_matrix / max(similarity_matrix))))



# Add color to the dendrogram branches based on clusters
clusters <- cutree(dendrogram, k = 3)  

# Define color palette for clusters
cluster_colors <- c("red", "black", "green")

# Assign colors to dendrogram branches
dendrogram_colored <- color_branches(dendrogram, k = 3, col = cluster_colors)

```

#### Hierarchical clustering across the different NSC and NP populations 
#### Figure 1 A

</br>

```{r Results dendrograms NSC NP, echo = FALSE}


# Plot the colored dendrogram
plot(dendrogram_colored, main = "Colored Dendrogram", ylab = "Distance")

```

</br>

#### Upset plots of shared gene signatures across the different studies 
#### Figure 1 B and 1 C and Supplementary Figure 1

```{r Upset plots, echo = TRUE}

#Load packages----
library(UpSetR)
library(ComplexHeatmap)
library(ggplot2)
library(ggpubr)
library(svglite)
library(ggplotify)
library(openxlsx)

#Load gene lists of NSC markers----
Hochgerner_RGL <- readLines("Hochgerner_RGL.txt")
Batiuk_AST4 <- readLines("Batiuk_AST4.txt")
Artegiani_NSC <- readLines("Artegiani_NSC.txt")
Bielefeld_A <- readLines("Bielefeld_A1.txt")
Harris_NSC <- readLines("Harris_NSC.txt")
Shin_NSC <- readLines("Shin_NSC.txt")
Wu_NSC <- readLines("Wu_qNSC.txt")

#UpsetPlot for common genes across NSC gene lists----
list1 <- list( Hochgerner_RGL,
               Batiuk_AST4,
               Artegiani_NSC,
               Bielefeld_A,
               Harris_NSC,
               Shin_NSC,
               Wu_NSC
               )
matrix1 <- list_to_matrix(list1)
colnames(matrix1) <- c("Hochgerner et al., 2020", "Batiuk et al., 2020" ,"Artegiani et al., 2017", "Bielefeld et al., 2024","Harris et al., 2021", "Shin et al., 2015", "Wu et al., 2025")

#In order to see how often each gene is present in the datasets
matrixSum1 <- matrix1
matrixSum1<- cbind(matrixSum1, Sum = rowSums(matrixSum1))
write.xlsx(matrixSum1, file = "2025-03-09_NSC_gene_signature.xlsx", rowNames = TRUE)

# Create the UpSet plot
m1 = make_comb_mat(matrix1)
graph1 <- UpSet(t(m1[comb_degree(m1) >= 4]),
    top_annotation = upset_top_annotation(t(m1[comb_degree(m1) >= 4]), add_numbers = TRUE),
    right_annotation = upset_right_annotation(t(m1[comb_degree(m1) >= 4]), add_numbers = TRUE),
    set_order = c("Hochgerner et al., 2020", "Batiuk et al., 2020" ,"Artegiani et al., 2017", "Bielefeld et al., 2024","Harris et al., 2021","Shin et al., 2015", "Wu et al., 2025"))


#Load gene lists of NP markers----
Hochgerner_nIPC <- readLines("Hochgerner_nIPC.txt")
Wu_nIPC <- readLines("Wu_nIPC.txt")
Bielefeld_RGL <- readLines("Bielefeld_RGL.txt")
Harris_NSC_IPCs <- readLines("Harris_NSC_IPCs.txt")
Bielefeld_NSC1 <- readLines("Bielefeld_NSC1.txt")
Bielefeld_NSC2 <- readLines("Bielefeld_NSC2.txt")

#Load gene lists of Neuroblast markers----
Hochgerner_Nb <- readLines("Hochgerner_Nb.txt")
Wu_Nb <- readLines("Wu_Nb.txt")
Artegiani_NP <- readLines("Artegiani_NP.txt")
Harris_Nb <- readLines("Harris_Nb.txt")


#UpsetPlot for common genes across Neuroblast gene lists----
list2 <- list( Hochgerner_Nb,
               Wu_Nb,
               Artegiani_NP,
               Harris_Nb
               )
matrix2 <- list_to_matrix(list2)
colnames(matrix2) <- c("Hochgerner et al., 2018","Wu et al., 2025", "Artegiani et al, 2017","Harris et al., 2021" )

#In order to see how often each gene is present in the datasets
matrixSum2 <- matrix2
matrixSum2<- cbind(matrixSum2, Sum = rowSums(matrixSum2))
write.xlsx(matrixSum2, file = "2025-02-21_Neuroblast_gene_signature.xlsx", rowNames = TRUE)

m2 = make_comb_mat(matrix2)
graph2 <- UpSet(t(m2[comb_degree(m2) >= 2]), 
                top_annotation = upset_top_annotation(t(m2[comb_degree(m2) >= 2]), add_numbers = TRUE),
                right_annotation = upset_right_annotation(t(m2[comb_degree(m2) >= 2]), add_numbers = TRUE),
                set_order = c("Hochgerner et al., 2018","Wu et al., 2025", "Artegiani et al, 2017","Harris et al., 2021"))

#UpsetPlot for common genes across NPs gene lists----
list3 <- list( Hochgerner_nIPC,
               Wu_nIPC,
               Bielefeld_RGL,
               Harris_NSC_IPCs,
               Bielefeld_NSC1,
               Bielefeld_NSC2
               )
matrix3 <- list_to_matrix(list3)
colnames(matrix3) <- c("Hochgerner et al., 2018","Wu et al., 2025", "Bielefeld et al., 2024 (RGL)","Harris et al., 2021","Bielefeld et al., 2024 (NSC1)","Bielefeld et al., 2024 (NSC2)" )

#In order to see how often each gene is present in the datasets
matrixSum3 <- matrix3
matrixSum3<- cbind(matrixSum3, Sum = rowSums(matrixSum3))
write.xlsx(matrixSum3, file = "2025-02-21_NP_gene_signature.xlsx", rowNames = TRUE)
write.xlsx(list(Sheet1 = matrixSum1, Sheet2 = matrixSum3, Sheet3 = matrixSum2), file = "2024-02-21_Gene_signatures.xlsx", rowNames = TRUE)

m3 = make_comb_mat(matrix3)
graph3 <- UpSet(t(m3[comb_degree(m3) >= 4]), 
                top_annotation = upset_top_annotation(t(m3[comb_degree(m3) >= 4]), add_numbers = TRUE),
                right_annotation = upset_right_annotation(t(m3[comb_degree(m3) >= 4]), add_numbers = TRUE),
                set_order = c("Hochgerner et al., 2018","Wu et al., 2025", "Bielefeld et al., 2024 (RGL)","Harris et al., 2021","Bielefeld et al., 2024 (NSC1)","Bielefeld et al., 2024 (NSC2)"))



```



```{r Upset plots results, echo=FALSE }
graph1_ggplot <- as.ggplot(graph1)
graph2_ggplot <- as.ggplot(graph2)
graph3_ggplot <- as.ggplot(graph3)
print(graph1_ggplot)
print(graph3_ggplot)
print(graph2_ggplot)



```
